Species richness destabilizes trophic communities in freshwater ecosystems

### 2020-03-19 manuscript/bef_stability/manuscript.Rmd
Add methods

---
title: Species richness destabilizes trophic communities in freshwater ecosystems
author: Alain Danet, Maud Mouchet, Elisa Thébault and Colin Fontaine
date: \today 
output:
  pdf_document:
    fig_caption: yes
fontsize: 12pt
header-includes:
   - \usepackage{setspace}
   - \doublespacing
geometry: margin=2cm
bibliography: references.bib
---

@@

- OFB
  - French long term protocol
  - standardized

- Electric fishing:
  - stuned fish
  - identification of the fishes
  - fish counted  
  - length of the fishes in mm by lot (4):
    - I: big fishes measured individually
    - S/L: fishes are grouped by species and by size 
      - 30 random fishes were measured in each lot
      - simulate the size of all the fishes of the lot:
	- normal distribution truncated at 95% (?)
	- mean and sd calculted from the sample of the lot 
    - G: little fishes are grouped by species
      - min and max size by lot 
      - simulate the size of all the fishes of the lot:
	- normal distribution truncated at 95% (?)
	- mean computed as ???  
	- se computed as ??? 
    - N:

- Protocols type:
  - complete
    - shallow river
  - partial:
    - deep and large river
    - partial over bank
    - partial by point
    - fusion of the two Protocols when "by point" all the point were done
      over bank

- Fishing operations and station
  - kept one protocol type by station
  - kept one sampling operation per year (of the season the most sampled in the
    time series)
  - remove operations which sampling effort (length of river sampled) superior to average
    plus or minus 30% of sd (by station)
  - kept station that had 10 or more samplings, i.e. followed 10 years or more

@@

- Diet data: 
  - 7? species over 57 had a piscivorous stage
  - 3? species had a strict piscivorous stage
  - ontogenic diet shift based on size

- 7 Resource nodes:
  - based on diet database
  - no information about their biomass

- Size class:
  - quantile: 9 size class per species

- Interaction:
  - Fish-resources and resources: based on diet data 
  - Fish-fish:
    1. diet data: is piscivorous?
    2. allometric relationship: predation range from .03 to .45 of body size

@@

- Biomass:
  - infered by allometric law 
  - approximated by $B = 0.01 \times (l^3.03)$

@@

- Collected during fishing operation:   
  - River size (width and depth)
  - Altitude and slope
  - distance to source
- Independant dataset from naiades: 
  - flow, temperature, DBO
    1. moving window average
    2. annual average 
    3. interpolation along rivers (OpenStars + SSN package)

@@

>>>>

---
title: Species richness destabilizes trophic communities in freshwater ecosystems
author: Alain Danet, Maud Mouchet, Elisa Thébault and Colin Fontaine
date: \today 
output:
  pdf_document:
    fig_caption: yes
fontsize: 12pt
header-includes:
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
geometry: margin=2cm
bibliography: references.bib
---

@@

---
#- OFB
  #- French long term protocol
  #- standardized
#- Electric fishing:
  #- stuned fish
  #- identification of the fishes
  #- fish counted  
  #- length of the fishes in mm by lot (4):
    #- I: big fishes measured individually
    #- S/L: fishes are grouped by species and by size 
      #- 30 random fishes were measured in each lot
      #- simulate the size of all the fishes of the lot:
	#- normal distribution truncated at 95% (?)
	#- mean and sd calculted from the sample of the lot 
    #- G: little fishes are grouped by species
      #- min and max size by lot 
      #- simulate the size of all the fishes of the lot:
	#- normal distribution truncated at 95% (?)
	#- mean computed as ???  
	#- se computed as ??? 
    #- N:
#- Protocols type:
  #- complete
    #- shallow river
  #- partial:
    #- deep and large river
    #- partial over bank
    #- partial by point
    #- fusion of the two Protocols when "by point" all the point were done
      #over bank
---

Fish communities were monitored across streams of France on the period
1995-2018 by a French governmental agency (OFB, previously AFB and ONEMA). We
kept only standardized protocols: complete and partial methods for respectively
little and big streams (detailed in Supplementary methods).  The sampling was
performed by electric fishing. All the stunned fish species were identified and
the individuals counted. Fish were classified in batch according to species and
size class. Depending on the size class, individual fishes were measured or
only a subsample. The size of individual fishes in subsampled batches were
inferred under normality distribution assumption (detailed in Supplementary
material).

@@

```{r}
myload(metaweb_analysis, dir= mypath("data"))
meta <- metaweb_analysis
#metaweb_analysis$resource
#meta$metaweb[meta$resource, meta$resource]

---
#- Diet data: 
  #- 7? species over 57 had a piscivorous stage
  #- 3? species had a strict piscivorous stage
  #- ontogenic diet shift based on size
#- 7 Resource nodes:
  #- based on diet database
  #- no information about their biomass
#- Size class:
  #- quantile: 9 size class per species
---

---
#- Interaction:
  #- Fish-resources and resources: based on diet data 
  #- Fish-fish:
    #1. diet data: is piscivorous?
    #2. allometric relationship: predation range from .03 to .45 of body size
---

Network inference was performed in two steps: determine trophic species,  build
a metaweb, subtract the metaweb according to the trophic species present in
community. Diet information were obtained from database and literature. Over
the 57 fish species, only seven had a piscivorous life stage and only three had
as strict piscivorous life stage. Fish species were largely omnivorous. Because
species have ontogenic diet shift, we created nine size classes by species, one
size class of one species constituting one trophic species. The bounds between
size classes were equidistant (percentile method). The metaweb described the
potential trophic interactions between the $`r 57*9`$ ($57\times9$) trophic species. 
We had seven resource nodes which were present in the diet of the trophic
species namely detritus, biofilm, phytoplankton, zooplankton, macrophages,
phytobenthos and zoobenthos. A trophic species belonged to a life stage if the
median of its size range fell in the range of the life stage. In turn, this
trophic species has feeding interactions with the resource nodes of its life stage.
Interactions between resource nodes are determined by literature. For example,
phytobenthos fed on detritus, while zooplankton fed on zooplankton and
phytoplankton. Fish-fish trophic interactions were determined by allometric
relationships as in many studies [@brose_predator_2019, @poisot_structure_2016,
@schneider_body_2012]. For a given trophic species, we first determined if its
size class end up in a life stage containing piscivorous diet. If it did, we
determined the prey size range of the trophic species from 3% to 45% of the
median of its size range. A feeding interaction was set for each trophic species
whose median of its size class fell into the prey range of the predator trophic
species. To sum up, the metaweb was obtained with allometric feeding
interactions (1) between trophic species, (2) between trophic species and
resource nodes and (3) between resource nodes.

The local networks (i.e. one network for each fish community sampling) was
obtained by determining the trophic species of each fish individual in the
community and interactions were set according to the metaweb. All the resource
nodes were assumed to be present in each local network. Finally, the
structure of the local networks varied according the presence of the trophic
species. The inferred interactions represent potential interactions (i.e.
not measured interactions) and are binary (i.e. presence or absence of feeding
interactions).

We computed the connectance ($C = L/N^2$, L and N being resp. the number of
links and of nodes) and the average trophic level of each network, the average
trophic level was weighted the biomass of each node (excluding resource nodes
for which we had no biomass) ($\overline{T} = \sum_i^N{B_i * T_i} \times
\sum_i^N{b_i}^{-1}$, $T_i$ and $B_i$ being resp. the trophic level and the
biomass of the node $i$). In order to have connectance at the species level
(i.e. not a trophic species level), we merge the trophic species of the network
by species and thereby aggregating the link by species.

@@

---
#- Biomass:
  #- infered by allometric law 
  #- approximated by $B = 0.01 \times (l^3.03)$
---

The body mass of each fish was approximated from their body size with an
allometric law ($B = 0.01 \times (l^3.03)$, $B$ being the body mass in grams and $l$ the
body size in millimeters). We have worked with body mass instead of abundances to avoid to sum
abundances of fishes that had really different sizes and so body masses. Body
mass are more meaningful in terms of ecosystem description. The biomass of each
trophic species (and species) was divided by the surface sampled (measured in
square meters) during the sampling event. The total biomass in a given
community are so expressed in $gm^{-2}$.

@@

---
#- Collected during fishing operation:
  #- River size (width and depth)
  #- Altitude and slope
  #- distance to source
#- Independant dataset from naiades: 
  #- flow, temperature, DBO
    #1. moving window average
    #2. annual average 
    #3. interpolation along rivers (OpenStars + SSN package)
---

Habitat description were obtained from fish sampling data. We kept variables
describing the ecosystem size namely the stream width and depth, the ecosystem
position (altitude and slope). Physico-chemical variables were obtained from
naiades (http://www.naiades.eaufrance.fr/)[http://www.naiades.eaufrance.fr/].
We chose flow and temperature as physical variables to describe streams as flow
impose swimming constraints which filter fish species that can establish [REF].
Water temperature is linked to physiological constraints namely thermal
tolerance and oxygen concentration tolerance [REF] since the level of water
saturation in oxygen is inversely linked to water temperature. We chose
Biochemical Oxygen Demand (BOD) for a proxy of enrichment in the ecosystem [REF
mccann]. BOD is also inversely linked to oxygen concentration in water.
The time resolution was heterogeneous according to the variable considered
(hourly, daily and monthly for temperature, flow and BOD). For each recording
station, we applied a moving window averaging (more details in supplementary
material) to get rid of seasonal effects and took the annual mean (i.e. one
value by year and by station. As the environmental recording stations were not
located at the same place than the fish stations, we interpolated data with
spatial statistical network model [@isaak_applications_2014] implemented in the 
`SSN` R package [@hoef_ssn:_2014]. The interpolation procedure is detailed
in supplementary material. For each fish station, environmental variables were
matched with the year of fish sampling (except for water temperature for which the 
record began in 2006). Then, the average temperature, flow, BDO, river width
and depth were computed as well as coefficient of variation ($CV = \sigma
\mu^{-1}$). As habitat and environmental variables are often colinear,
a principal component analysis (PCA) was performed on the scaled variables. The
PCA explained % of the variance. An axis rotation was performed to maximize the
representation of the variables by the principal components
(Fig. \@ref{fig:pca-rotated}). We kept five axis, representing respectively the
(1) river size (river width & depth), (2) the altitude, slope and decreasing
average temperature, (3) average enrichment (BDO) and CV of flow, (4) river
size CV (CV of river width and depth) and (5) CV of enrichment (CV of BDO).
Additionally, we logged the axis one, two and three to normalize their
distribution and we reversed the third axis to get increasing average
temperature with increasing values of the axis.

@@

---
#- Fishing operations and station
  #- kept one protocol type by station
  #- kept one sampling operation per year (of the season the most sampled in the
    #time series)
  #- remove operations which sampling effort (length of river sampled) superior to average
    #plus or minus 30% of sd (by station)
  #- kept station that had 10 or more samplings, i.e. followed 10 years or more
---

Only one protocol
type was kept by station. We kept only one sampling event by year to homogenize
the dataset [numbers: month of sampling, number of days bw samplings] (detailed
in Supplementary material).
